Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-24 doc/ApTranscriptome_manuscript.Rmd
FINAL edits to ms

@@

**Affilitiations**

\textsuperscript{1} Department of Biology, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{2} Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, Tenessee 37996 \newline
\textsuperscript{3} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{4} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
\textsuperscript{5} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
\textsuperscript{6} Center for Macroecology, Evolution and Climate, University of Copenhagen

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis* across 12 temperature intervals that spanned their entire thermal breadth. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptation. To better understand the relationship between these two mechanisms, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found substantial differences in their patterns of gene expression. *Aphaenogaster picea* responded to extreme temperatures primarily through upregulation, with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures is not just a more extreme version of the acclimation response, but instead involves a qualitative shift in mechanisms from those used to cope with rarely encountered stressors.

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.4°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced responses to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_slim_2014-10-22.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)

@@

## Reaction norms of thermally-responsive transcripts differ between species

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these 9,052 transcripts, 80% (...) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.

@@

![Deviations from expected numbers of transcripts in matched observation data for transcript expression type data. The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit). Shown are **Low**, **High** and **Bimodal** expression types for (A) *A. picea* and (B) *A. carolinensis* against all expression types in the opposite species. The dashed boxes indicate the diagonals (e.g. **High** - **High**).](../results/matched_observations_Fig1.png)

The molecular processes in the set of transcripts that switched from upregulation in *A. picea* to downregulation in *A. carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism. In addition, there were GO terms related to repair (e.g. histone modification), transcript regulation (e.g. positive regulation of cellular process) and cellular organization (e.g. transport, organelle organization). Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of transcript downregulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures. Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately ... at the higher temperature extremes and a shift of ...  at low temperature extremes. Again, there was greater inertia of gene expression only in response to high temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures the species did not differ in mean temperature of downregulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

@@

We tested the *genetic assimilation hypothesis*, a shift from inducible to constitutive expression of stress-response genes, by comparing the log ratios of relative inducibility to relative constitutive expression. Specifically, this hypothesis predicted a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme; transcripts upregulated at high temperatures in *A. picea* (Fig. 3A) or transcripts upregulated at low temperatures in *A. carolinensis* (Fig. 3B). In fact, for both comparisons we found a weakly positive relationship.

![Scatterplots of (A) log ratio of inducibility in *A. picea* over inducibility in *A. carolinensis* against log ration of constitutive expression in *A. carolinensis* over constitutive expression in *A. picea* for transcripts with **High** expression in *A. picea*, and (B) log ratio of inducibility in *A. carolinensis* over inducibility in *A. picea* against log ration of constitutive expression in *A. picea* over constitutive expression in *A. carolinensis*. Blue lines and confidence intervals are from are ordinary least squares regressions.](../results/genetic_assimilation_Fig3.png)

In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A. carolinensis*.

@@

In contrast to cold tolerance, the enhanced upper thermal limit in *A. picea* is best explained by the *tolerance* hypothesis. High temperatures were associated with significantly fewer upregulated transcripts in *A. carolinensis*, and a large proportion (361 of 1,238 transcripts) of those transcripts upregulated in *A. picea* were either downregulated or expressed at negligible levels overall in *A. carolinensis*. Similarly, significantly higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 2B-D). This difference between species is also evident among candidate genes known to be involved in stress response. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include constitutive defenses [@barshis2013],  enhanced proteome stability [@ghosh2010] or behavioral changes [@sunday2014] to tolerate thermal stress. These differences are in line with expectations that *A. carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts. This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al. [-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations. Both the genetic assimilation and tolerance routes routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair. Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.

Given the differences in the patterns of thermal responsiveness between species (Fig. 1, 2), it is worth noting a number of similarities. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Table 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012]. The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role. Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temperature are bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig. 2). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify transcripts that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation. Thus, our method provides an example of how RNA-seq data can be used to understand gain meaningful insight on the process of evolution.

@@

Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions [@hoekstra2013].  Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al. 2001).  Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.  Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.

@@

## Thermally-responsive transcripts

>>>>

**Short title**: Thermal reactionomes of cool and warm-climate ant species

@@

**Affiliations**

\textsuperscript{1} Department of Biology, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{2} Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, Tenessee 37996 \newline
\textsuperscript{3} Vermont Genetics Network, University of Vermont, Burlington, Vermont 05405 \newline
\textsuperscript{4} Department of Biological Sciences, North Carolina State University, Raleigh, North Carolina 27695 \newline
\textsuperscript{5} Harvard Forest, Harvard University, Petersham, Massachusetts 01336 \newline
\textsuperscript{6} Center for Macroecology, Evolution and Climate, University of Copenhagen, Universitetsparken 15, DK-2100 Copenhagen

**Corresponding author**

John Stanton-Geddes

Department of Biology, University of Vermont, 109 Carrigan Dr, Burlington, VT 05405.

Phone: 802-656-2922

Email: john.stantongeddes.research@uvm.edu

**Keywords**

*Aphaenogaster*, gene expression, plasticity, reactionome, transcriptome

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperature intervals that spanned their entire thermal breadth. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

Coping with progressive changes in climate requires both acclimation and long-term adaptation. To better understand the relationship between these mechanisms, we compared gene expression as a function of temperature in two ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. We found substantial differences in their patterns of gene expression. Aphaenogaster picea responded to extreme temperatures primarily through upregulation, with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated metabolic genes. These results suggest that adaptation to chronic high temperatures is not just a more extreme version of the acclimation response, but instead involves a shift in mechanisms from those used to cope with rarely encountered stressors.

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences warmer (mean annual temperature (MAT) at collection site = 14.6°C) and less variable (temperature seasonality = 7,678) temperatures than *A. picea* (MAT = 4.6°C, seasonality = 10,008; [@hijmans2005]), with correspondingly higher critical maximum temperatures (44.7°C for *A. carolinensis* versus 41.3°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*) allowing us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated a *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense hypothesis* [@hofmann1996; @feder1999; @kultz2005] proposes that species extend their thermal limits through a stronger induced response to provide greater protection from more frequently encountered stressors. This hypothesis would predict that the cool-adapted *A. picea* would induce more genes in response to low temperatures than would the warm-adapted *A. carolinensis*, which would induce more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should also be stronger for more commonly experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adapations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Finally, the *genetic assimiliation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation.

@@

```{r load, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#options(scipen = 1, digits = 3)
load("ms_slim_2014-10-22.RData")
library(data.table)
library(ggplot2)
library(reshape2)
library(pander)
library(RColorBrewer)
library(topGO)
library(plyr)

@@

## Reaction norms of thermally-responsive transcripts

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of the Formicidae hits, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these 9,052 transcripts, ~80% (7,201) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate of 8% is likely a lower bound for the true proportion of genes that are thermally responsive.

@@

![Deviations from expected numbers of transcripts in matched observations of transcript expression type. The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit). The dashed boxes indicate the diagonals (e.g. **High** - **High**). (A) **Low** and **Bimodal** expression types for *A. carolinensis* showing bias away from all other categories in *A. picea*. (B) **High** and **Bimodal** expression types for *A. picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**). The complete set of matched observations is shown in Fig. S1.](../results/matched_observations_Fig1.png)

The molecular processes in the set of transcripts that switched from upregulation in *A. picea* to downregulation in *A. carolinensis* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` significantly enriched GO terms (Table S1), of which `r round((length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term))) / (nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea))*100, 0)`% (`r (length(grep("metab", Ar.int.A22.low.gsea$Term)) + length(grep("metab", Ar.int.A22.bim.gsea$Term)))`) were related to metabolism. In addition, there were GO terms related to repair (e.g. histone modification), transcript regulation (e.g. positive regulation of cellular process) and cellular organization (e.g. transport, organelle organization). Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism whereas the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes. There was greater inertia of gene expression in response to high and low temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures **Intermediate** transcripts from *A. picea* were downregulated at higher temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)`) than those from *A. carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

@@

We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the midpoint temperature (19.5°C). If stress-response genes have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme; transcripts upregulated at high temperatures in *A. picea* (Fig. 3A) or transcripts upregulated at low temperatures in *A. carolinensis* (Fig. 3B). In fact, for both comparisons we found a weakly positive relationship.

![Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A. picea* (*P* < 0.001, r^2^ = 0.07) and (B) **Low** transcripts in *A. carolinensis* (*P* < 0.001, r^2^ = 0.1). Blue lines and confidence intervals are from ordinary least squares regressions. ](../results/genetic_assimilation_Fig3.png)

In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in far-right column of Fig. 1A). Rather than having constitutively elevated expression as predicted, these genes were uniformly low or not expressed (mean TPM = 0.08) in *A. carolinensis*.

@@

In contrast to cold tolerance, the enhanced upper thermal limit in *A. picea* is best explained by the *tolerance* hypothesis. High temperatures were associated with significantly fewer upregulated transcripts in *A. carolinensis*, and a large proportion (361 of 1,238 transcripts) of those transcripts upregulated in *A. picea* were either downregulated or expressed at negligible levels overall in *A. carolinensis*. Similarly, significantly higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* towards the high thermal limit (Fig. 2B-D). This difference between species is also evident among candidate genes known to be involved in stress response. Of the `r nrow(hsp_all)` transcripts annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to thermal extremes in *A. picea* whereas only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*. Altogether, these results suggest that mechanisms other than the heat shock response are acting to maintain protein stability in face of temperature increases, which may include novel constitutive defenses [@barshis2013; @meyer2011; @oneil2014], enhanced proteome stability [@ghosh2010] or behavioral quiescence [@sunday2014] to tolerate thermal stress. These differences are in line with expectations that *A. carolinensis*, with a growing season over twice the length of its northern congener, may be better able to afford to restrain from foraging in suboptimal conditions. In fact, quiescence under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts. This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013], although in a short-term selection experiment for heat tolerance, Sikkink et al. [-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations. Both the genetic assimilation and tolerance routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair. Whether a particular taxon evolves one strategy over another may be related to the intensity, frequency and duration of temperature stress in a given environment.

Given the differences in the patterns of thermal responsiveness between species (Fig. 1, 2), it is worth noting a number of similarities. In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures  (Table 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. Among closely-related *Drosophila* species, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. To the extent that the number of thermally responsive genes is related to additive genetic variation, our results also suggest that adaptability to high temperatures is more genetically limited than low temperatures, perhaps placing these northern species at risk of competitive replacement by more heat-tolerant southern species [@rodriguez-cabal2012]. The two species also showed similar numbers of genes with a **Bimodal** response (Table 1), indicating a generalized stress-response role. Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression.  Only 15-20% of transcripts upregulated in response to temperature were bimodal, and although upper activation thresholds differed significantly between species, lower thresholds were virtually identical (Fig. 2). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

A major contribution of this study is the construction of a reactionome for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. While we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.

@@

Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions [@hoekstra2013].  Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins [@krebs1997] or by increasing the thermal stability of proteins at the expense of catalytic activity [@arnold2001].  Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.  Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.

@@

## Identification of thermally-responsive transcripts

We quantified expression of each transcript using the program `Sailfish` [@patro2014] and use the bias-corrected transcripts per million (TPM) [@wagner2012] as our measure of transcript expression. We included the contaminant transcripts identified by DeconSeq at the quantification stage to avoid incorrectly assigning reads to other transcripts, but removed these from further analyses. As preliminary examination of the data (Supplemental Methods 1) indicated that the 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental Methods 1).
